---
title: "Statistical Inference Course Project Part 2"
author: "Calvin Seto"
date: "September 7, 2015"
output: html_document
---

## Synopsis
In this report, we'll analyze a data set describing the length of teeth (measured in microns) in guinea pigs who received Vitamin C using two different methods (orange juice or ascorbic acid) and three different dosage levels (0.5, 1, and 2 mg).

## Data Exploration
Figure 1 shows a conditioning plot of guinea pig teeth length versus dose, given the type of supplement.  We see
1-using supp OJ or VC, the length increases as does increases from 0.5 to 1.0 to 2.0
2-under OJ, the length range is about 12 to 27, and under VC, the length range is about 7 to 27
3-for both OJ and VC, the change in length is about the same for doses 0.5 and 1.0
under VC, increasing the dose from 1.0 to 2.0 shows a bigger increase in length than for OJ

We'll create the following subsets
ssOJ - teeth length of guinea pigs who received orange juice supplements
ssVC - teeth length of guinea pigs who received ascorbic acid supplements
ss0510 - teeth length of guinea pigs who received dosages of 0.5 and 1 mg.
ss0520 - teeth length of guinea pigs who received dosages of 0.5 and 2 mg.
ss1020 - teeth length of guinea pigs who received dosages of 1 and 2 mg.

```{r}
require(graphics)
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

```{r}
ssOJ <- subset(ToothGrowth, supp == "OJ")
ssVC <- subset(ToothGrowth, supp == "VC")
ss0510 <- subset(ToothGrowth, dose == 0.5 | dose == 1)
ss0520 <- subset(ToothGrowth, dose == 0.5 | dose == 2)
ss1020 <- subset(ToothGrowth, dose == 1 | dose == 2)
```

range(ssOJ$len)
range(ssVC$len)
range(ss0510$len)
range(ss0520$len)
range(ss1020$len)


## Confidence Intervals and tests

## Results

## Conclusions



population of 60 guinea pigs, divided into 6 groups of 10

group 1
group 2
group 3
group 4
group 5
group 6

6 different regimes of vitamin C
regime 1 OJ 0.5
regime 2 OJ 1.0
regime 3 OJ 2.0
regime 4 VC 0.5
regime 5 VC 1.0
regime 6 VC 2.0

after 42 days, pigs were euthanized and the mean length of teeth for each group were measured

ToothGrowth dataset
60 obs of 3 variables-len, supp, dose

row 1 is mean length of tooth for guinea pig 1 from group 1
row 2 is mean length of tooth for guinea pig 2 from group 1
row 3 is mean length of tooth for guinea pig 3 from group 1
...
row 11 is mean length of tooth for guinea pig 11 from group 2
row 12 is mean length of tooth for guinea pig 12 from group 2
row 13 is mean length of tooth for guinea pig 13 from group 2
...
row 60 is mean length of tooth for guinea pig 60 from group 6

1-compare by supplement orange juice using 3 dosages vs chemical using 3 dosages
2-compare by dosage 0.5 using 2 supplements vs 1.0 using 2 supplements vs 2.0 using 2 supplements

so they measured each day and the data set contains the mean of 42 values?
