---
title: "Statistical Inference Course Project Part 1"
author: "Calvin Seto"
date: "September 7, 2015"
output: html_document
---

## Overview
The exponential distribution describes the time between events in a Poisson process, or a process in which events occur continuously and independently at a constant average rate.  But in this exercise, we will investigate the distribution of averages of 40 exponentials and use the Central Limit Theorem to explain how it looks bell shaped.

The Central Limit Theorem states that the distribution of averages of independent and identically distributed variables becomes that of a standard normal (mean = 0, standard deviation = 1) as the sample size increases.

The Law of Large Numbers says that the average limits to what it's estimating, the population mean.

## Simulations
We'll use simulations to generate a lot of exponentials and plot the means and show the resulting distributions look normal.
We set $$\lambda = 0.2$$ for all simulations.

```{r}
set.seed(168)
library(ggplot2)
nosim <- 1000
lambda <- 0.2
mu <- 1/lambda
s <- 1/lambda
cfunc <- function(x, n) sqrt(n) * (mean(x) - mu) / s
m1 <- matrix(sample(rexp(40,0.2), nosim * 10, replace = TRUE), nosim) # 10x1000 matrix sample 10,000 from 40 random exponentials with replacement
m2 <- matrix(sample(rexp(40,0.2), nosim * 20, replace = TRUE), nosim) # 20x1000 matrix sample 20,000 from 40 random exponentials with replacement
m3 <- matrix(sample(rexp(40,0.2), nosim * 30, replace = TRUE), nosim) # 30x1000 matrix sample 30,000 from 40 random exponentials with replacement
r1 <- apply(m1, 1, cfunc, 10) # apply CLT approximation for each row (sample) of 10 passing n=10 
r2 <- apply(m2, 1, cfunc, 20) # apply CLT approximation for each row (sample) of 20 passing n=20
r3 <- apply(m3, 1, cfunc, 30) # apply CLT approximation for each row (sample) of 30 passing n=30
dat <- data.frame(
    x = c(r1, r2, r3),
    size = factor(rep(c(10, 20, 30), rep(nosim, 3))))

```


## Sample Mean versus Theoretical Mean
```{r}
1/0.2
```

## Sample Variance versus Theoretical Variance
```{r}
1/0.2
```

## Distribution
```{r}

```
